---
author: "Thalia"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(dplyr)
library(gplots)
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

### Data Types and Structures in R
#### 1. Data types

Everything in R is an object. There are mainly five **data types**: 

```{r, echo=FALSE}
data_types <- c("Character", "Integer", "Numeric", "Logical", 'Complex')
description <- c('Strings', 'Intergers', 'Integers & Factions',
                 'Boolean', 'Complex value')
examples <- c("\"Hello\", \"2.3 \", \"TRUE\"", 2, 12.3, "TRUE & FALSE", "3+2i")

tabl <- data.frame(data_types, description, examples)
kable(tabl, format= "html") %>%
  kable_styling("striped", full_width = FALSE) %>%
  column_spec(1, width = "10em", bold = TRUE) %>%
  column_spec(2, width = "10em") %>%
  column_spec(3, width = "13em") 
```

R provides many functions to examine features of vectors and other objects, for example:

```{r, echo=FALSE}
functions <- c("class()", "typeof()", "length()", "attributes()")
description <- c("The object's datatype (high-level)",
                 "The object's datatype (lowlevel)",
                 "The length of the object", 
                 "The names of the metadata of the object")

tabl <- data.frame(functions, description)
kable(tabl, format= "html") %>%
  kable_styling("striped", full_width = FALSE) %>%
  column_spec(1, width = "10em", bold = TRUE) %>%
  column_spec(2, width = "20em") 
```

| Expression     | Result     |
|:-----------------|:-------------| 
|  class(10)      |   `r class(10)`     |
|  typeof(10)     |   `r typeof(10)`    |
|  class(7.12)    |   `r class(7.12)`   |
|  typeof(7.12)   |   `r typeof(7.123)` |
|  class("HELLO") |   `r class("HELLO")`|
|  length("HELLO")|   `r length("HELLO")`|
|  class("TRUE")  |   `r class("TRUE")` |
|  class(TRUE)    |   `r class(TRUE)`   |
|  class(4+2i)    |   `r class(4+2i) `  |
|  class(NA)      |   `r class(NA)`     |
|  length(NA)     |   `r length(NA)`|
|  class(NULL)    |   `r class(NULL)`   |
|  length(NULL)    |   `r length(NULL)`   |


**Note**: there are multiple classes that are grouped together as "numeric" classes. The 2 most common of which are double and integer. Most math is done using double precision, so that is often the default storage.

#### 2. Data structures

Elements of these data types may be combined to form **data structures**:

- Vectors
- Lists
- Matrices 
- Arrays 
- Factors
- Data Frames

**I. Definition of vectors**

Vector is a basic data structure in R. It contains element of the \textcolor{red}{same} type. The data types can be logical, integer, double, character, complex or raw. There are different ways to define vectors:

**1. c() function**
```{r}
a <- c(1,2,3,4)
print(a)
```
**2. Colon operator**
```{r}
b <- 1:10
print(b)
```
**3. Sequence operator**
```{r}
c <- seq(5, 9, by = 2)
print(c)
```
**4. Replicate Elements Of Vectors or Lists**
```{r}
d <- rep('hi', 5)
print(d)
```
**Assign name of elements in list**
```{r}
e <- c('apple' = 1, banana = 2, "kiwi fruit" = 3, 4)
print(e)
```

**II. Definition of lists**

Lists are the R objects which contain elements of **different** types like − numbers, strings, vectors and another list inside it
```{r}
# Defination of Vectors
lst <- list('Thalia', 26, 'Seattle')
print(lst)
```

**Assign name of elements in list**
```{r}
lst <- list('Thalia', 26, 'Seattle')
names(lst) <- c('Name', 'Age', 'Home')
print(lst)
```

**III. Definition of metrices**

Matrices are the R objects in which the elements are arranged in a two-dimensional rectangular layout. They contain elements of the **same** types.
```{r}
mtx1 <- matrix(c(1:8), nrow=4, byrow=TRUE)
print(mtx1)
```

**Assign name of row and columns in matrix**
```{r}
# Define the column and row names
rownames = c("row1", "row2", "row3", "row4")
colnames = c("col1", "col2")

# Create the matrix
mtx2 <-  matrix(c(1:8), nrow = 4, byrow = FALSE, 
                dimnames = list(rownames, colnames))
print(mtx2)
```

**IV. Definition of arrays**

Arrays are the R data objects which can store data in more than two dimensions. For example − If we create an array of dimension (2, 3, 4) then it creates 4 matrices each with 2 rows and 3 columns. Arrays can store only **same** data type.

```{r}
vector1 <- c(5, 9, 3)
vector2 <- c(10, 11, 12, 13, 14, 15)
# You can use the c() function to combine vectors 
print(c(vector1, vector2))
# Take these vectors as input to the array.
result1 <- array(c(vector1, vector2), dim = c(3,3,2))
# Create the matrix.
print(result1)
```
**Assign name of rows, columns, and matrix in arrays**
```{r}
# You can also assign name to each dimension of the array
col_names <- c("Col1", "Col2", "Col3")
row_names <- c("Row1", "Row2", "Row3")
matrix_names <- c("Matrix1", "Matrix2")

result2 <- array(c(vector1, vector2), dim = c(3,3,2), 
                 dimnames=list(row_names, 
                               col_names,
                               matrix_names))
print(result2)

```

**V. Definition of Factors**

Categorical variable where each level is a category. It can store both strings and integers.

```{r}
# Factors are created using the factor () function by taking a vector as input.
loc <- c("East","West","East","North","North","East","West","South","West","East","North")
print(factor(loc))

val <- c(1, 2, 3, 2, 2, 2, 3, 1)
print(factor(val))
```

**VI. Definition of Data Frame**

A data frame is a table or a two-dimensional array-like structure in which each column contains values of one variable and each row contains one set of values from each column. The data stored in a data frame can be of numeric, factor or character type.

```{r}
employee_df <- data.frame(
   emp_id = c (1:5), 
   emp_name = c("Thalia", "Hung", "Tony", "Ayla", "Jimmy"),
   salary = c(623.3, 515.2, 611.0, 729.0, 843.25), 
   start_date = as.Date(c("2012-01-01", "2013-09-23", "2014-11-15", "2014-05-11",
      "2015-03-27")),
   stringsAsFactors = FALSE
)
# Print the data frame.			
print(employee_df) 
```
**Rename the columns in dataframe**
```{r}
colnames(employee_df) <- c('ID', 'NAME', 'PAY', 'DATE')
print(employee_df)

```

| Expression     | Result     |
|:-----------------|:-------------| 
|  class(list(1,2,3,4,'hello'))      |   `r class(list(1,2,3,4,'hello'))`     |
|  typeof(list(1,2,3,4,'hello'))     |   `r typeof(list(1,2,3,4,'hello'))`    |
|  length(list(1,2,3,4))      |   `r length(list(1,2,3,4))`     |
|  typeof(list(1,2,3,4)     |   `r typeof(list(1,2,3,4))`    |
|  class(c(1,2,3,4))    |   `r class(c(1,2,3,4))`   |
|  typeof(c(1,2,3,4))   |   `r typeof(c(1,2,3,4))` |
|  class(c('a', 'b', 'c', 'd')) |   `r class(c('a', 'b', 'c', 'd'))`|
|  typeof(c('a', 'b', 'c', 'd')) |   `r typeof(c('a', 'b', 'c', 'd'))`|
|  class(factor(c(1,2,2,3)))  |   `r class(factor(c(1,2,2,3)))` |
|  typeof(factor(c(1,2,2,3)))  |   `r typeof(factor(c(1,2,2,3)))` |
|  class(factor(c("Female","Male")))  |   `r class(factor(c("Female","Male")))` |
|  typeof(factor(c("Female","Male")))  |   `r typeof(factor(c("Female","Male")))` |

- Note how class() gives you high-level data type while typeof() gives you low-level data type.
- Note factor is stored as integer at the lower level no matter whether it is defined using the vecotors of integer.

$ \text{This is the END}$ 
$ \text{This is the END}$ 